The imprint of global magma oceans on exoplanet demographics

### 2022-11-23 src/tex/ms.tex
prepare description of planetary sample

\subsubsection{Stellar luminosity tracks}
\begin{note}
    Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
    Since the emergence of a runaway greenhouse phase on a planet is highly dependent on the level of radiation it receives, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
    Stellar ages are notoriously poor-constrained (CITE); we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
    We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.
    Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.

\begin{figure}[ht!]
    \script{plot_luminosity_tracks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/luminosity_tracks.pdf}
        \caption{
            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models in \citet{Baraffe1998}.
            Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
        }
        \label{fig:luminosity_tracks}
    \end{centering}
\end{figure}

\subsubsection{Orbital and bulk properties of the planet sample}

\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions. Probably replace SAG13-based content below.}
\begin{note}
    The NASA Exoplanet Program Analysis Group chartered Science Analysis Group 13 (SAG 13)...
    The inferred occurrence rate density can be expressed as a power law in planet radius and period
    \begin{equation}
        \frac{\partial^2n}{\partial \log R \, \partial \log P} = \Gamma R^{\alpha} P^{\beta}
    \end{equation}
that is broken in planet radius, resulting in the free parameters $\Gamma$, $\alpha$, $\beta$, and a breakpoint radius $R_\mathrm{break}$.
\end{note}

@@

\subsubsection{Transit probability}
\todo[inline]{describe how we model inclinations and transits of synthetic planets}
\begin{note}
    Not all planets are transiting from our point of view.
    We model the occurrence of transits by assuming random, isotropic orientations of planetary orbits and calculating the transit impact parameter $b = \cos(i)/R_\star$ for each planet.
    Only planets with $|b| < 1$ are transiting and further considered.
    For these cases we calculate the transit depth
    \begin{equation}\label{eq:transitdepth}
        \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2
    \end{equation}\todo[inline]{check if transit depth or duration are actually used further down}
    and transit duration
    \begin{equation}\label{eq:transitduration}
        t_{\mathrm{T}} = \frac{R_\star P}{\pi a} \sqrt{1 - b^2},
    \end{equation}
    which are relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
    After excluding all non-transiting planets, the size of our nominal sample of potentially observable planets drops to $N = \Nplanets$.
\end{note}

@@

\subsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?}
\begin{note}
Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the \textit{Kepler} mission~\citep{Fulton2017,VanEylen2018}.
    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}

>>>>

\subsubsection{Stellar luminosity tracks}
    Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
    Since the emergence of a runaway greenhouse phase on a planet is highly dependent on the level of radiation it receives, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
    Stellar ages are notoriously poor-constrained (CITE); we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
    We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.
    Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
\begin{figure}[ht!]
    \script{plot_luminosity_tracks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/luminosity_tracks.pdf}
        \caption{
            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models in \citet{Baraffe1998}.
            Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
        }
        \label{fig:luminosity_tracks}
    \end{centering}
\end{figure}

\subsubsection{Orbital and bulk properties of the planet sample}
Next, we assigned planetary systems to the synthetic stellar sample.
The orbital parameters and bulk properties of these simulated planets are informed by statistics from the \kepler mission; we adopted the planetary occurrence rates recently derived in \citep{Bergsten2022}.
Following \citep{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
\begin{equation}
    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
\end{equation}
where $F_0$ ...
\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}

@@

\subsubsection{Transit probability}
\begin{note}
    Not all planets are transiting from our point of view.
    We model the occurrence of transits by assuming random, isotropic orientations of planetary orbits and calculating the transit impact parameter $b = \cos(i)/R_\star$ for each planet.
    Only planets with $|b| < 1$ are potentially observable in transit and further considered.
    For these cases we calculate the transit depth
    \begin{equation}\label{eq:transitdepth}
        \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2
    \end{equation}\todo[inline]{check if transit depth or duration are actually used further down}
    and transit duration
    \begin{equation}\label{eq:transitduration}
        t_{\mathrm{T}} = \frac{R_\star P}{\pi a} \sqrt{1 - b^2},
    \end{equation}
    which are relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
    Excluding all non-transiting planets diminishes the sample to \SI{\percentageTransiting}{\percent} of its original size.
\end{note}

@@

\subsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?}
\begin{note}
Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}
